Gap formation and stability in non-isothermal protoplanetary discs

### 2014-11-19 paper/results.tex
updating results section - MKL

According to \cite{li00}, increasing $h$ increases linear growth rates
of the RWI because it is pressure-driven. However, in the case 
of disc-planet interaction, increasing $h$ has a stabilizing effect
through the gap profile because it results in smoother gap
edges. The fact that we observe smaller growth rates as $h$ is
increased indicates that for planetary gaps, the importance of $h$ on
the \emph{linear} RWI is through the setting the gap profile, i.e. basic
state for the instability (as opposed to the linear response).

@@

\subsection{Non-linear evolution without planet
  influence} \label{nonlinearplanetoff} 
{\bf typical rossby numbers for these `planet-off' vortices? do
  stronger vortices (beta=0.1) dissipate faster? if we want to say the
  vortices dissipate viscously, then need to quote a viscosity
  (e.g. reynolds stress) and compute a viscous time and compare to the
  observed decay time. Or attribute to numerical viscosity. It might
  be simplest to just say the vortices decay without the planet. This
  section is mainly used to contrast with planet-on simulations, so
  don't need to go into detail. 
}

@@

Fig.\ref{planetofflifetimeplot} shows evolution of the $m=1$ surface
density amplitude, which represents the post-merger single vortex. For
completeness we also ran intermediate cases with $\tilde{\beta}=0.5$
and $5.0$. The vortices simply decay, on a timescale of $O(10^3)$
orbits. We will see in the that section that this behaviour is very
different to when the planet potential is kept on. Interestingly,
though, vortices arising from the stronger instabilities (shorter
cooling times or lower $\tilde{\beta}$) decays faster.

@@

We now examine long-term simulations of gap-edge vortices for
$\tilde{\beta}=0.1,0.5,1,5,10$. The planet potential is kept on
throughout.  
% up to a total of
%$2000P_0$. 
%Planets were left free to interact dynamiclly with the disk
%and vortices after $t=30P_0$ as apposed to previous section.  
We employ a grid with $(N_r,N_{\phi})=(512,1024)$ in order for these
simulations to be computationally feasible. We also use a larger
disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimize boundary
effects on vortex evolution, and apply an open boundaries at
$r=r_\mathrm{in},\,r_\mathrm{out}$. We comment that lower-resolution
simulations with $(N_r,N_{\phi})=(256,512)$ were initially carried
out, which show similar behaviour and trends as the high-resolution
runs reported below.

@@

\subsection{Generic evolution} 
The linear growth of the RWI and vortex-formation is followed by 
vortex merging. We now find merging timescales independent of
$\tilde\beta$, and by $60P_o$ only one vortex remains. 
%as the growth of these modes are now
%accelerated by planetary influence as apposed to `planet off' simulations.
% i.e. only the $m=1$ mode remains. 
The evolution of the amplitude of the $m=1$ surface density mode,
averaged over $r-r_p\in[2,10]r_h$ is shown in Fig~\ref{lifetimeplot} 
for different $\tilde\beta$. 
{\bf any particular reason for the larger radial range for average?
  did we get bigger vortices?} 
% {\bf should state (also for planet-off runs)
% what the radial range is used for averaging the Fourier amplitudes}

@@

In all cases the system remains in a quasi-steady state for
$\gtrsim800P_0$ with a single vortex circulating 
the outer gap edge at the local Keplerian  
frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
surface density perturbation in this state. During this stage, the 
vortex intensifies. As the vortex forms it has a
Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
indicating anti-cyclonic
motion. As the $m=1$ amplitude grows the Rossby number also increases
in magnitude, 
%relative spin of the vortex
%continuously grows along with it
, reaching a characteristic value of $Ro\approx-0.4$ for all cooling
rates. {\bf is the the value reached just before decay for all cooling?}

Surface density pertibations relative to $t=0$
at the center of vortices can reach values of $\Delta\Sigma/\Sigma_0=7$ 
for all cases of $\tilde\beta$ in this quasi-steady state and
can be found as high as $\Delta\Sigma/\Sigma_0= 9$ times for the 
$\tilde\beta=10$ case. 
The gap profiles formed also apply a surface density boost in the vortex 
region as the vortices are located at the gap edge maxima.
The average value of relative density pertibation in the outer gap region of 
$r=r_{p}+2.5r_h$
has a value $\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$ for all times and
cooling rates.
Thus the surface density increases caused by gap formation are small 
compared to the max amplitudes of the vortices.
Vortex growth is attributed to
the continuous generation of vorticity by the planet-disc interactions. 
This is supported by the evidence that in long term simulations where the
planet is removed, such as seen in Fig \ref{planetofflifetimeplot},
the amplitude of
the vortices are not seen to grow without the planet influence.

@@

Interestingly, in most cases we observe a sudden drop in 
the $m=1$ amplitude within the simulation time, which occurs over
timescales of $t\in[20P_0,70P_0]$, i.e. much shorter than the duration of the
quasi-steady state. Dissipation timescales do not correlate with $\tilde\beta$
 but instead are paramtrized by vortex Mach numbers.
%{\bf is this dissipation timescale independent of cooling? Trend is based on mach values which dont follow linear trend } 
%Eventually
%the $m=1$ amplitudes are also characterized by 
After these characteristic drops, the instability is never seen to
reform at such large overdensities and surface density bumps are never larger
than that due to the planetary wakes. We designate the time of these
amplitude drops as the lifetime 
of the quasi-steady vortex.
After the vortices dissipate the RWI is not seen to be excited again and no
vortices are seen to refrom in the simulations time.
%The instability is not seen to emerge again after the intial vortex dissipation
% due to the smoothening out of the gap edge during dissipation by the
% transfer of density from the once tightly held vortex region to the surrounding disk and gap.  
{\bf Discussed in next section:  question: does instability
  actually happen again? if it does, but just weaker, that may be
  because the first vortex has smoothed out the gap edge, so
  instabilities following that would be weaker.}

\begin{figure}
  \includegraphics[width=\linewidth,height=\linewidth]{figures/vortex2D}
  \caption{Relative surface density perturbation for the
    $\tilde\beta=0.1$ case during quasi-steady state with a single
    vortex. The plot for other values of the cooling time $\tilde{\beta}$ are
    similar.% {\bf is this plot truncated? in the text it says the
      %domain goes to $r=45$?}
    %The large scale non-axisyemetric $m=1$ overdensity can be
    %seen. 
    \label{Vortex2D}} 
\end{figure}

{\bf
\subsection{Additional analysis on vortex dissipation}
Conjecture: vortex death due to vortex shocking the system and
dissipating its energy. discuss one case. here we want to explain or
suggest why vortices die} 
%Rossby values have dynamics that are coupled with the lifetime and
%$m=1$ amplitude of the vortices.

\begin{figure}
\subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock1}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock2}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock3}
  } \\[-0.98cm]
\subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock4}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock5}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock6}
  } 
  \caption{Relative density pertibation of vortices (top) and density gradient
 (bottom) for before (left), during (middle) and after dissipation (right).
 Large density gradients are characteristics of shocks and become prominant
 around vortex during dissipation} \label{shockplot}
\end{figure}

Gradients of surface density are seen to
spike up in wake-like features around the vortex during dissipation. 
Snapshots of relative density pertibation and density gradients during the
quasi steady state, during disspation and just before death are displayed in
Fig. \ref{shockplot} for $\tilde\beta=1.0$. Quasi-steady vortices have large 
elliptical shapes, while dissipating vortices are denser and smaller in spacial
extent.
As the vortices grow in intensity wakes are seen to extend from them. These
wakes correlate with large gradients in surface density and are first seen
when vortices are in the later half of their quasi steady states. During this
phase the average value of the gradients along the shock region are
$|\nabla\Sigma| \approx 4 \Sigma$. Just before the vortices start dissipation
this quantity sharply increases to a value of $ \approx6 \Sigma$ in which it
 remains for rest of vortex lifetime.  
As the vortex dies off as described above its associated Rossby number quickly
approaches $0$ indicating a halting of cyclonic motion.

In addition, we also observed large increases in the Mach number near
the vortex during dissipation, which indicate susceptibility to
shocks. Mach numbers with repect to the vortex bulk motion around the vortex
 region of $|r-r_{vortex}|<2H$ {\bf (somewhat arbitrary, large enough to have bulk of vortex mach number increases quickly radially away from vortex)}
%, where $H$ denotes the gas scale height, 
are shown in Figure \ref{machplot} for various cooling rates over time.
% $\approx 0.8$ after the intial formation of the vortices for all $\tilde\beta$. 
During the quasi-steady state the Mach number increases at a
steady rate with larger growth rate for higher $\tilde\beta$.
In all cases Mach numbers are seen to spike up at a time which correlates with
the start of vortex dissipation.
% Time values in
% which Mach numbers reach values of 1 approximately correlate with the start 
%of dissipation and during dissipation values quickly increase to larger max values.
The timescales of vortex dissipation also correlates with 
Mach numbers reached during dissipation. The quickest dissipation occurs for 
$\tilde\beta=0.5$ which is on order of $20P_0$ and reaches Mach number
 $\approx1.4$ while the $\tilde\beta=0.1$ case which dissipates on scales o
f $90P_0$ reaches Mach numbers $\approx1.2$.
Thus, the sudden drop in vortex amplitude may  
be a result of the development of associated spiral shocks dissipating
its energy \citep[cf.][who 
  suggested damping due to planet-induced shocks]{fu14}.

{\bf note that fu et al suggested vortex damping by the planet
  shocks. here, we are suggesting vortex dissipates energy by shocking
  the disk. our evidence is that we observe vortex growth before it
  dissipates, and spiral shocks associated with the vortex always
  develop before vortex death. once it starts shocking the disc it
  loses energy. if the planet-induced shocks were responsible, death
  should be more gradual. can we think of other evidences that it's
  not the planet shock that's responsible?}
{\bf show
  this. e.g. vortex before, during and after death. maybe show
  snapshots of surface density perturbation (vortex) compared with
  $\nabla\Sigma/\Sigma$ (vortex shocks). demonstrate vortex shape
  changes and shocks develop prior ot death. state whether this is
  generic for all coolings}. 
{\bf show mach number plot, or quote values. relevant mach
  number is fluid flow speed with respect to 
  vortex bulk motion around the disc. does the typical mach number
  near the vortex just before its death depend on cooling?}

Vortex growth during the quasi-steady state and dissipation also acts to smooth
out the gap edge. To characterize the gap, we azimuthally and radially average
the relative density pertibation within $r\in[r_{planet},r_{out}]$ where
 $r_{out}=$ is defined as the outer gap radius defined as point where
$\langle\Delta \Sigma(r_{out})/\Sigma(t=0)\rangle_{\phi}=0$. Thus a more 
negative gap depth parameter correlates with a steeper gap edge. In
 Fig. \ref{gapdepth} the gap depth parameter with respect to simulation time
for $\tilde\beta=1.0$ is shown. The gap depth is seen to slowly increase as
vortex reaches its max amplutude at $t=1000P_0$ which is followed by a sharp
increase which correlates with the time of vortex dissipation.
%Of note is also that the large overdensity the vortex
%creates during it's quasi-steady state distorts the background
%disk.  
The smoothening out of the gap edge during later half of quasi steady state
 is because the vortex has an associated viscosity $O(10^{-2})$.
%{\bf quote a typical value of alpha viscosity associated with vortex}.
 This acts
against gap-opening by the planet, so that the condition for the RWI
becomes less favourable as the vortex grows and hence is why vortices are not
seen to form again in simulation timescales.
The sudden smoothening of gap egde during dissipation is due to the large
influx of density from the vortex to the gap.
{\bf not sure what you mean by self-limiting here:self-limiting
  behaviour?} 
{\bf show this (smoothening of gap edge due to vortex growth),
  e.g. general vortensity evolution or surface density 
  evolution}

\begin{figure}
    \includegraphics[width=\linewidth]{figures/mach}
 \caption{Running plot of Mach number relative to vortex within $2H$ of the
 vortex. Mach numbers increase at rates correlated with cooling rates than
 spike up at times coinciding with the start of vortex dissipation as seen in
 Fig \ref{lifetimeplot}} \label{machplot}
\end{figure}

\begin{figure}
    \includegraphics[width=\linewidth]{figures/gapdepth}
 \caption{Running average of relative density pertibation within the outer half
 of gap} \label{gapdepth}
\end{figure}

@@

{\bf describe dependency on
cooling. Do we have enough data to have a `lifetime v.s. beta'
  plot? here we want to explain why vortex lives longer with longer
  cooling, but very long cooling results in shorter lifetimes. 
  Consider the following quantities as a function of cooling:
  initial strength of the single vortex; growth rate of single vortex
  strength from its formation to just before death; ampiltude of
  vortex just before death; disc temperature in the vortex region.

  we expect the initial $m=1$ vortex (after merging) to be weaker with
  increasing cooling time because the instability is weaker. (see
  meheut et al 2013, mnras, 430, 1988 who show that the initial vortex
  amplitude after linear growth is correlated with the linear growth
  rate).

  suppose the vortex shocks when it reaches a certain amplitude (which
  probably increases with longer cooling rate because it's more
  difficult to shock a hotter disc). as we
  increase the cooling time, the initial $m=1$ vortex is weaker, it
  also grows more slowly because gap-opening is opposed by a hotter
  disc. the above implies that it takes longer for the vortex reach
  the shock-inducing amplitude and dissipate.

  but the above is not consistent with non-monotonic behaviour, which
  predicts the $\tilde{\beta}=10$ vortex to survive longest. this is
  because we expect the initial vortex to be weakest (check) which
  means it should take the longest to grow and shock. check
  whether there's anything special about this case. one possibility is
  that vortex growth during the quasi-steady state is actually faster
  because of the raised temperature (li et al 2000 show increased
  temperature favours RWI - possibiliy evidenced here by a stronger
  vortex before death than other cases)

  perhaps have a case of $\tilde{\beta}=7.5$ as further evidence of
  non-monotonic behaviour
}

\begin{figure}
    \includegraphics[width=\linewidth]{figures/betaplot}
 \caption{Lifetime of vortices with repect to cooling paramter $\tilde\beta$ for both resolution of $(N_r,N_\phi)=(256,512)$ and $(512,1024)${\bf should I only show higher resolution points?} } \label{betaplot}
\end{figure}

@@

These results were reproduced in lower resolution simulations as well. Vortex
shocks were similarly seen during vortex dissipation and non-monotonic
 dependence on lifetime with $\tilde\beta$ was similarly found.
Vortex lifetimes are considerably longer in these runs with the maximum
 lifetime longer than the simulation timescale of $2000P_o$ for 
$\tilde\beta=5.0$ which again has $h\approx0.06$.
This inconsistancy in timescales, may be due to resolution as a finer 
grid spacing is required to accurately
reproduce jump conditions along shocks and which cause vortex dissipation.
 {\bf this is the best reason I could personally think of, for lower beta runs vortices last slightly longer but for higher values it is quite drasticly different. Not sure how numerical dissipation as you mentioned would be a factor in this case}
%be the optimum lifetime for long-term simulations of
%isothermal disks with varying intial temperature profiles $h$.

>>>>

According to \cite{li00}, increasing $h$ increases linear growth rates
of the RWI because it is pressure-driven. However, in the case 
of disc-planet interaction, increasing $h$ has a stabilizing effect
through the gap profile because it results in smoother gap
edges. The fact that we observe smaller growth rates as $h$ is
increased indicates that for planetary gaps, the importance of $h$ on
the \emph{linear} RWI is through setting up the gap profile, i.e. basic
state for the instability (as opposed to the linear response).

@@

\subsection{Long term evolution} \label{nonlinearplanetoff} 
{\bf typical rossby numbers for these `planet-off' vortices? do
  stronger vortices (beta=0.1) dissipate faster? if we want to say the
  vortices dissipate viscously, then need to quote a viscosity
  (e.g. reynolds stress) and compute a viscous time and compare to the
  observed decay time. Or attribute to numerical viscosity. It might
  be simplest to just say the vortices decay without the planet. This
  section is mainly used to contrast with planet-on simulations, so
  don't need to go into detail. 
}

@@

Fig.\ref{planetofflifetimeplot} shows evolution of the $m=1$ surface
density amplitude, which represents the post-merger single vortex. For
completeness we also ran intermediate cases with $\tilde{\beta}=0.5$
and $5.0$. The vortices simply decay, on a timescale of $O(10^3)$
orbits. We will see in the next section that this behaviour is very
different to when the planet potential is kept on. Interestingly,
though, vortices arising from the stronger instabilities (shorter
cooling times or lower $\tilde{\beta}$) decays faster.

@@

We now examine long-term simulations of gap-edge vortices for
$\tilde{\beta}=0.1,0.5,1,5,10$. The planet potential is kept on
throughout.  
% up to a total of
%$2000P_0$. 
%Planets were left free to interact dynamiclly with the disk
%and vortices after $t=30P_0$ as apposed to previous section.  
We employ a grid with $(N_r,N_{\phi})=(512,1024)$ in order for these
simulations to be computationally feasible. We also use a larger
disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimize boundary
effects on vortex evolution, and apply open boundaries at
$r=r_\mathrm{in},\,r_\mathrm{out}$. We comment that lower-resolution
simulations with $(N_r,N_{\phi})=(256,512)$ were initially carried
out, which show similar behaviour and trends as the high-resolution
runs reported below.

@@

\subsection{Generic evolution} 
The linear growth of the RWI and vortex-formation is followed by 
vortex merging. We now find merging timescales independent of
$\tilde\beta$, and by $60P_o$ only one vortex remains.  
%as the growth of these modes are now 
%accelerated by planetary influence as apposed to `planet off' simulations.
% i.e. only the $m=1$ mode remains. 
The evolution of the amplitude of the $m=1$ surface density component,
averaged over $r-r_p\in[2,10]r_h$ is shown in Fig~\ref{lifetimeplot} 
for different $\tilde\beta$. 
{\bf any particular reason for the larger radial range for average?
  did we get bigger vortices?} 
% {\bf should state (also for planet-off runs)
% what the radial range is used for averaging the Fourier amplitudes}

@@

In all cases the system remains in a quasi-steady state for
$\gtrsim800P_0$ with a single vortex circulating 
the outer gap edge at the local Keplerian  
frequency. Fig. \ref{Vortex2D} shows a typical plot of the relative 
surface density perturbation in this state. During this stage, the 
vortex intensifies. As the vortex forms it has an initial 
Rossby number $Ro\approx-0.1$ for all $\tilde\beta$
(indicating anti-cyclonic
motion). As the $m=1$ amplitude grows the Rossby number also increases
in magnitude, 
%relative spin of the vortex
%continuously grows along with it
, reaching a characteristic value of $Ro\approx-0.4$ for all cooling
rates. {\bf is the the value reached just before decay for all cooling?}

The surface density perturbation (relative to $t=0$)  
can reach values of $\Delta\Sigma/\Sigma_0 \sim 7$ at the center of vortices
for all cases of $\tilde\beta$ in quasi-steady state, and
was found as high as $\Delta\Sigma/\Sigma_0 \sim 9$ for the 
$\tilde\beta=10$ case. This large increase in surface density is due 
vortex growth, since there is continuous generation of vorticity by
planet-disc interaction. This is supported by the observation that in
the previous simulations without the planet, the amplitude of
the post-merger vortex does not grow (Fig
\ref{planetofflifetimeplot}).  We remark that the increase in the
vortex surface density directly due to the pile-up of material at the
gap edge (because of gap-opening) is small: the average surface density
perturbation near the outer gap edge, $r\sim r_{p}+2.5r_h$, is 
$\langle\Delta\Sigma/\Sigma_0\rangle_\phi<1$.

@@

Interestingly, in most cases we observe a sudden drop in 
the $m=1$ amplitude within the simulation time, which occurs over
timescale between $20P_0$ to $70P_0$, i.e. much shorter than the duration of the
quasi-steady state. We did not find this dissipation timescale to
correlate with $\tilde\beta$. %but instead are paramtrized by vortex Mach numbers.
%{\bf is this dissipation timescale independent of cooling? Trend is
%based on mach values which dont follow linear trend }  
%Eventually
%the $m=1$ amplitudes are also characterized by 
%After these characteristic drops in vortex amplitude, the instability
%is never seen to reform at such large overdensities: the vortex surface
%densities are never larger than that due to the planetary wakes. 
We designate the time at which these amplitude drops as the lifetime 
of the quasi-steady vortex. After the vortices decay the RWI is 
not seen to be excited again and no vortices are seen to refrom within the
simulations time. 
{\bf Perhaps a snapshot of what the gap edge looks like after vortex death.
}
%The instability is not seen to emerge again after the intial vortex dissipation
% due to the smoothening out of the gap edge during dissipation by the
% transfer of density from the once tightly held vortex region to the surrounding disk and gap.  
% {\bf Discussed in next section:  question: does instability
%   actually happen again? if it does, but just weaker, that may be
%   because the first vortex has smoothed out the gap edge, so
%   instabilities following that would be weaker.}

\begin{figure}
  \includegraphics[width=\linewidth,height=\linewidth]{figures/vortex2D}
  \caption{Relative surface density perturbation for the
    $\tilde\beta=0.1$ case during quasi-steady state with a single
    vortex. The plots for other values of the cooling time
    $\tilde{\beta}$ are similar.
    % {\bf is this plot truncated? in the text it says the
    % domain goes to $r=45$?}
    % The large scale non-axisyemetric $m=1$ overdensity can be
    % seen. 
    \label{Vortex2D}} 
\end{figure}

\subsection{Additional analysis on vortex dissipation}
{\bf are these high resolution results? vortex lifetime in snapshots
  not consistent with amplitude plots}
%Conjecture: vortex death due to vortex shocking the system and
%dissipating its energy. discuss one case. here we want to explain or
%suggest why vortices die} 
%Rossby values have dynamics that are coupled with the lifetime and
%$m=1$ amplitude of the vortices.
In this subsection we examine the sudden vortex decay observed in our
simulations in more detail. Fig. \ref{shockplot} show snapshots
of the vortex leading up to its decay for the case $\tilde{\beta}=1$. 
The surface density perturbation and the surface density gradient are
shown. In quasi-steady ($t=700P_0$) the vortex is elongated, but becomes
more compact as it grows before decaying. {\bf rough estimate of
  vortex aspect-ratio in the three snapshots? 
  (length/width, by inspection is ok)}

Notice in Fig. \ref{shockplot} the appearence of wakes extending from
either side of the vortex at $t=1300P_0$ as it grows. These 
wakes correlate with large gradients in surface density (bottom
panel), and are first seen in the later half of the quasi-steady state.       
% As the vortices grow in intensity wakes are seen to extend from them. 
During this phase the average value of the surface density gradient
along the wakes is $|\nabla\Sigma/\Sigma| \sim 4 $.{\bf quote something
  dimensionless, or say `in code units'}  
Just before the vortices start to decay, we observe this quantity
sharply increases to $ \sim 6 $, and remains around this value
until the vortex dies out, at which point the 
associated Rossby number quickly approaches zero.

\begin{figure}
\subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock1}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock2}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock3}
  } \\[-0.98cm]
\subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock4}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock5}
  }
\hfill
  \subfigure{
    \includegraphics[width=0.3\linewidth]{figures/shock6}
  } 
  \caption{Evolution of the vortex in the case $\tilde{\beta}=1$
    just before its sudden decay. The surface density perturbation
    (top) and the associated surface density gradient (bottom) are shown.
    \label{shockplot}}
  % Large surface density gradients become prominant
  % around vortex during dissipation} 
\end{figure}

We also measured large increases in the Mach number near
the vortex when it begins to decay. Figure \ref{machplot} plots the
Mach number $M=|\bm{v} - \bm{v}_\mathrm{vor}|/c_s$, where
$\bm{v}_\mathrm{vor}$ corresponds to the bulk velocity of the vortex
around the disc. {\bf check - i assume this is the definition of mach
  number used here}. Values in Fig. \ref{machplot} have been averaged over
a region within $2H$ of the vortex centre.    
% Mach numbers with repect to the vortex bulk motion around the vortex
% region of $|r-r_{vortex}|<2H$
% are shown in  for various cooling rates over time.
During the quasi-steady state the Mach number increase
steadily, and for all cases $M$ reaches a maximum at 
% with larger growth rate for higher $\tilde\beta$.
a time coincident with the start of vortex decay.

Putting the above observations together, we suggest the sudden decay
in the vortex amplitude is due to shock formation \emph{by the
  vortex}. When the vortex reaches large amplitude, it begins
to induce shocks in the surrounding fluid, as supported by the
increase in Mach number and the appearence of wakes with large surface
density gradients. The vortex may then lose energy through shock
dissipation. In addition, a strong vortex (or shock formation) 
also smoothes out the gap structure that originally gave rise to the
RWI, this would oppose vortex growth. We examine this below.

\begin{figure}
    \includegraphics[width=\linewidth]{figures/mach}
 \caption{Running-time average {\bf ?} of the Mach number relative to
   the vortex, averaged over a region within $2H$ of the  
   vortex centre. This plot can be compared to the evolution of the
   vortex amplitude shown in Fig. \ref{lifetimeplot}.
   {\bf remember to update plot to include full simulation duration}
   \label{machplot}}
\end{figure}

\subsection{Effect of vortex dissipation on the gap structure} 
{\bf are these high resolution results?}
We find vortex growth during the quasi-steady state and its dissipation
also modifies the gap structure. Here, we 
define a gap depth parameter by taking the azimuthal and radial average
of the relative surface density perturbation within $r\in[r_p,r_e]$ where
$r_{e}$ is defined as the outer gap radius such that 
$\langle\Delta \Sigma(r_e)/\Sigma(t=0)\rangle_{\phi}=0$. We are
interested in the region $r \in[r_p,r_e]$ since the vortex is
located at the outer gap edge. A more negative gap depth parameter
correlates with a steeper gap edge. 
{\bf notation clash: rout already used as outer disc boundary - choose
different notation. updated to $r_e$ for now}

Fig. \ref{gapdepth} shows the evolution of the gap depth parameter
for $\tilde\beta=1.0$. The magnitude of the gap depth is seen to 
slowly increase as vortex reaches its maximum amplutude at
$t=1000P_0$, followed by a sharp decrease coincident with vortex
dissipation.

We see that the dissipation of the vortex has a `gap-filling'
effect, and we find that the gap edge is smoothed during this
process. This can be interpreted as the vortex providing a
viscosity, and we measure a typical alpha viscosity $\alpha = O(10^{-2})$
associated with the vortex. This acts against gap-opening
by the planet, and   
%The smoothening out of the gap edge during later half of quasi steady state
%is because the vortex has an associated viscosity $\alpha \sim O(10^{-2})$.
the condition for the RWI
becomes less favourable, which explains why vortices do not
reform again (at least within the simulation timescale).

%The sudden smoothening of gap egde during dissipation is due to the large
%influx of density from the vortex to the gap.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gapdepth}
  \caption{Running-time average of the gap depth parameter, defined as
    the relative surface density perturbation averaged over the
    outer half of the gap.} \label{gapdepth}
\end{figure}

@@

% {\bf describe dependency on
% cooling. Do we have enough data to have a `lifetime v.s. beta'
%   plot? here we want to explain why vortex lives longer with longer
%   cooling, but very long cooling results in shorter lifetimes. 
%   Consider the following quantities as a function of cooling:
%   initial strength of the single vortex; growth rate of single vortex
%   strength from its formation to just before death; ampiltude of
%   vortex just before death; disc temperature in the vortex region.

%   we expect the initial $m=1$ vortex (after merging) to be weaker with
%   increasing cooling time because the instability is weaker. (see
%   meheut et al 2013, mnras, 430, 1988 who show that the initial vortex
%   amplitude after linear growth is correlated with the linear growth
%   rate).

%   suppose the vortex shocks when it reaches a certain amplitude (which
%   probably increases with longer cooling rate because it's more
%   difficult to shock a hotter disc). as we
%   increase the cooling time, the initial $m=1$ vortex is weaker, it
%   also grows more slowly because gap-opening is opposed by a hotter
%   disc. the above implies that it takes longer for the vortex reach
%   the shock-inducing amplitude and dissipate.

%   but the above is not consistent with non-monotonic behaviour, which
%   predicts the $\tilde{\beta}=10$ vortex to survive longest. this is
%   because we expect the initial vortex to be weakest (check) which
%   means it should take the longest to grow and shock. check
%   whether there's anything special about this case. one possibility is
%   that vortex growth during the quasi-steady state is actually faster
%   because of the raised temperature (li et al 2000 show increased
%   temperature favours RWI - possibiliy evidenced here by a stronger
%   vortex before death than other cases)

%   perhaps have a case of $\tilde{\beta}=7.5$ as further evidence of
%   non-monotonic behaviour
% }

\begin{figure}
    \includegraphics[width=\linewidth]{figures/betaplot}
 \caption{Lifetime of vortices with repect to cooling paramter $\tilde\beta$ for both resolution of $(N_r,N_\phi)=(256,512)$ and $(512,1024)${\bf update to high resolution results} } \label{betaplot}
\end{figure}

@@

These results were reproduced in lower resolution simulations as well. Vortex
shocks were similarly seen during vortex dissipation and non-monotonic
 dependence on lifetime with $\tilde\beta$ was similarly found.
Vortex lifetimes are considerably longer in these runs with the maximum
 lifetime longer than the simulation timescale of $2000P_o$ for 
$\tilde\beta=5.0$ which again has $h\approx0.06$.
This inconsistancy in timescales, may be due to resolution as a finer 
grid spacing is required to accurately
reproduce jump conditions along shocks and which cause vortex dissipation.
 {\bf this is the best reason I could personally think of, for lower
   beta runs vortices last slightly longer but for higher values it is
   quite drasticly different. Not sure how numerical dissipation as
   you mentioned would be a factor in this case}

%be the optimum lifetime for long-term simulations of
%isothermal disks with varying intial temperature profiles $h$.
